Introduction to\\ Swedish Civil Contingencies Agency (MSB) methodological support for introducing \ac{isms}.

### 2019-03-06 msbframework/msbintro/msbintro.tex
Removed two slides, lecture to big
Started to update the lecture slides, to be up to date with the latest changes from MSB.
Placed all figures in a separate folder

% $Id$
%\documentclass[handout]{beamer}
\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage{lmodern,textcomp}
\usepackage[T1]{fontenc}
\usepackage[swedish,english]{babel}
\usepackage{url}
\usepackage{graphicx}
\usepackage{color}
\usepackage{subfig}
\usepackage{csquotes}
\usepackage{acro}
\usepackage[backend=biber,style=alphabetic,maxbibnames=99]{biblatex}
\addbibresource{msbintro.bib}
\DeclareAcronym{isms}
{
  short = ISMS,
  long = Information Security Management Systems,
}

\setbeamertemplate{bibliography item}[text]

\mode<presentation>{%
  \usetheme{Frankfurt}
  \setbeamercovered{transparent}
  \usecolortheme{seagull}
}
\setbeamertemplate{footline}{\insertframenumber}

\title[Intro MSB]{%
  Introduction to\\
  Swedish Civil Contingencies Agency (MSB) methodological support for introducing \ac{isms}.
}
\author{Carina Bengtsson, Daniel Bosk and Lennart Franked\footnote{%
  Detta verk är tillgängliggjort under licensen Creative Commons 
  Erkännande-DelaLika 2.5 Sverige (CC BY-SA 2.5 SE).
	För att se en sammanfattning och kopia av licenstexten besök URL 
	\url{http://creativecommons.org/licenses/by-sa/2.5/se/}.
}}
\institute[MIUN IST]{%
  Department of Informationsystem and Technologies (IST),\\
  Mid Sweden University, Sundsvall.

  %Avdelningen för informations- och kommunikationssytem (IKS),\\
  %Mittuniversitetet, Sundsvall.
}
\date{\today}

%\pgfdeclareimage[height=0.65cm]{university-logo}{MU_logotyp_int_CMYK.pdf}
%\logo{\pgfuseimage{university-logo}}

@@

% Since this a solution template for a generic talk, very little can
% be said about how it should be structured. However, the talk length
% of between 15min and 45min and the theme suggest that you stick to
% the following rules:

% - Exactly two or three sections (other than the summary).
% - At *most* three subsections per section.
% - Talk about 30s to 2min per frame. So there should be between about
%   15 and 30 frames, all told.

@@

\subsection{MSB}

\begin{frame}{MSB}
  \begin{itemize}
    \item Swedish Civil Contingencies Agency (MSB).
    \item \enquote{MSB is responsible for issues concerning civil protection,
        public safety, emergency management and civil defence, as long as no
        other authority has responsibility\cite[About MSB]{msbse}\@.
        .}
    \item This includes protection of our information systems.
  \end{itemize}
\end{frame}

%\begin{frame}{MSB:s organisation}
%  \begin{figure}
%    \includegraphics[height=0.7\textheight]{msb-organisation.png}
%    \caption{MSB:s organisation.}
%  \end{figure}
%\end{frame}

\begin{frame}[allowframebreaks]{Tieto-breakdown}
  \cite{Lindkvist2012tdf} gives the following summary:
  \begin{description}
    \item[Friday afternoon] Tieto notices a disruption in their IT-systems.
      350 pharmacies lost contact with their IT-systems.
      Many larger organisations are also affected, amongst other a larger
      logistical company.

    \item[Sunday afternoon] Tieto is reporting hardware malfunction, and start
     the necessary steps to fix the malfunction.

    \item[Monday morning] The logistical company are unable to handle its
      operation, and cannot reach its employees.
      The vehicle inspection agency are unable to access their IT-system. Since
      they were handling over 20\, 000 vehicle inspections a day, it might
      result to a driving ban for some vehicles, since they cannot report
      approved inspections.
      Nacka municipality, have to resort to Facebook and Twitter for
      communicating within the municipality.

    \item[Monday afternoon] Social office in Nacka and Sollentuna are unable to
      pay child support.
      Stockholm City absence reporting system for the schools are down.

    \item[Wednesday lunch] All the pharmacies have gotten access to their
      IT-systems again.

    \item[11 days] The logistical company can start using their IT-system.
      The organisation where still recovering from the disruption, two months
      after.
  \end{description}
\end{frame}

\subsection{Methodological support}

\begin{frame}{Informationssäkerhet.se}
  \begin{itemize}
    \item In 2011 MSB ran a project called SVISA\@: 'Stöd för Verksamheters 
      InformationsSäkerhetsArbete'.

    \item Resulted in \url{informationssäkerhet.se}.

    \item Is meant to give practical advice for systematically incorporate
      information security into an organisation.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Methodological overview}
  \begin{figure}
    \includegraphics[width=6cm]{metodstod-overview.png}
    \caption{Overview over the methodological support.}
  \end{figure}
\end{frame}

\section{Prepare}

\subsection{Introduction}

\begin{frame}{What is information security?}
  \begin{itemize}
    \item Occurs together with other processes and organisations.

    \item Information security in an organisation does not have a value by
      itself. It needs to be integrated into the organisation to be effective.

\begin{frame}{What is information security?}
  \begin{itemize}
    \item Ability to preserve the requirements and expectations that exist on
      information in an organisation.

    \item Amongst other to protect towards disruptions, such as what happened to
      Tieto.

\begin{frame}{Demands and expectations on information}
  \begin{description}
    \item[Traceability] Who have taken part of, or changed the information?
    \item[Non Repudiation] It should not be possible to deny an act.
    \item[Authentication] Establish an entities identity.
    \item[Authorization] To give an authenticated entity certain permissions.
  \end{description}
  These will be covered later in the course.
\end{frame}

\begin{frame}{Why protect the information?}{Non-mandatory}
  \enquote{Good for business}
  \begin{description}
    \item[Reputation:] Who will let a company handle their information, if the
      company is known for treating their data carelessly.

    \item[Financial:] Strong reputation is better for the economy, and the cost
      of dealing with security incidents will be less.

    \item[Internal efficiency:] No loss of information or disruptions in the
      work.

    \item[Quality:] This will hopefully lead to a increase in work quality.

\begin{frame}{Why protect the information?}{Mandatory}
  \begin{description}
    \item[Personal Data Act 1998:204] adds restrictions on how an organisation manages
      personal data.

    \item[GDPR] General Data Protection Regulation, Applies in all EU country
      starting from 25th of May 2018.
    \item[Public Access to Information and Secrecy Act 2009:40] Says that certain
      information \emph{must} be available for the public, while other information
      should \emph{not} be.

    \item[The Archives Act 1990:782] says that the government needs to archive
      all public documents.

    \item[MSBFS 2016:1] applies to governmental agency and their
      work with information security.

\begin{frame}{GDPR}{General Data Protection Regulation}
  \begin{itemize}
    \item Replacing Personal Data Act 1998:204
    \item Regulates how personal data are allowed to be stored and used.
      \begin{itemize}
        \item Protects a persons right of privacy\cite[art. 5]{eu:gdpr}
        \item Personal data can only be stored for as long as there is a need
          for it\cite[art. 5]{eu:gdpr}.
        \item Each person must give explicit consent to a company/organisation
          to store, publish or share personal data about them\cite[art.
          7]{eu:gdpr}.  
        \item Ensure that a person have the right to get access
          to all personal
          data that has been collected about them\cite[art. 15]{eu:gdpr}.
        \item Persons right to erasure\cite[art. 17]{eu:gdpr}. 
      \end{itemize}
  \end{itemize}
\end{frame}
\begin{frame}{GDPR cont.}{General Data Protection Regulation}
    \begin{itemize}
      \item Possibility to transfer personal data (data
          portabililty)\cite[art. 20]{eu:gdpr}.  
      \item Data protection by design and default \cite[art. 25]{eu:gdpr}.
      \item Records regarding processing activities must be maintained \cite[art.
        30]{eu:gdpr}.
      \item Personal data breaches must be reported to the supervisory
        authority, without undue delay, not later than 72 hours\cite[art.
        33]{eu:gdpr}.
    \end{itemize}
\end{frame}
\begin{frame}{GDPR cont.}{General Data Protection Regulation}
\begin{itemize}
    \item Individuals needs to be informed about the personal data
        breach\cite[art. 34]{eu:gdpr}.
    \item An independent agency should ensure that these rules are followed
      \cite[art 51]{eu:gdpr}.
    \item If the organisation does not comply to these regulations, they can be
      fined up to 4\% worldwide turnover or €20 Million, whichever is greater
      \cite[art 83]{eu:gdpr}.
    \item and much more.
\end{itemize}
\end{frame}
\begin{frame}{MSBFS 2016:1}
  \begin{itemize}
    \item Due to increased electronic information exchange in the society, there
      is now demands put on how governmental agency work with information
      security.

    \item The code of statutes came into effect 1th of February 2010.

\begin{frame}{MSBFS 2016:1}
  1 §  Denna författning innehåller föreskrifter som ansluter till
  bestämmelserna om statliga myndigheters informationssäkerhet i 19§
  förordningen (2015:1052) om krisberedskap och bevakningsansvariga
  myndigheters åtgärder vid höjd beredskap.

@@

\end{frame}
\begin{frame}{MSBFS:2016:1}
  Detta innebär bland annat att en myndighet måste: 
  \begin{enumerate}
    \item upprätta en informationssäkerhetspolicy och andra styrande dokument 
      som behövs för myndighetens informationssäkerhet,
    \item utse en eller flera personer som leder och samordnar arbetet med 
      informationssäkerhet,
    \item klassificera sin information med utgångspunkt i krav på 
      konfidentialitet, riktighet och tillgänglighet,
    \item utifrån risk- och sårbarhetsanalyser och inträffade incidenter 
      avgöra hur risker ska hanteras, samt besl uta om åtgärder för 
      myndighetens informationssäkerhet,
    \item dokumentera granskningar och säkerhetsåtgärder av större betydelse 
      som har vidtagits.
  \end{enumerate}
\end{frame}

@@

\begin{frame}{ISO 27000}
  \begin{itemize}
    \item This is not just for governmental agencies.

    \item Any organisation can certify themselves for ISO27000

    \item MSB:s methodological support is adapted to the international
      standards.

\begin{frame}{Governance documents}
  \begin{itemize}
    \item Plan of action, policies, guidelines, \dots

    \item Policy: What do we want to achieve?
      One extensive, or multiple smaller ones.

    \item The governance documents should be integrated into the organisations
      current structure.

\begin{frame}{Information security in different levels}
  \begin{figure}
    \includegraphics[width=\textwidth]{infosak-levels.png}
    \caption{Different levels where information security can be conducted.}
  \end{figure}
\end{frame}

\begin{frame}{Purpose with \ac{isms}}
  \begin{figure}
    \includegraphics[width=\textwidth]{infosak-pdca.png}
    \caption{To convert requirements to an active protection.}
  \end{figure}
\end{frame}

\begin{frame}{ISO 27001}
  \begin{itemize}
    \item MSB's methodological support, describes how to build an \ac{isms},
      based on ISO 27001.

    \item ISO 27001 is a continuous process that strives towards constant
      improvements on how to work and use different security solutions in
      regards to information security.

    \item It is important to adapt this after the organisation, but this does
      not mean that parts can be skipped.

\begin{frame}{PDCA}
  \begin{description}
    \item[Plan] Plan, analyze and design.
    \item[Do] Establish.
    \item[Check] Follow up.
    \item[Act] Improve.
  \end{description}
  Then start over again.
\end{frame}

\begin{frame}{ISO 27002 -- What should be done?}
  All of this is covered in different chapters in ISO 27002. We will cover parts
  of this during the next lecture.
\end{frame}

\subsection{Committed Management}

\begin{frame}{Create commitment}
  \begin{itemize}
    \item The top management have the overall responsibility for the
      organisation -- This involves information security and security incidents.

    \item Important to make the management understand the importance of
      information security.

\begin{frame}{Motivation}
  \begin{itemize}
    \item What positive effects might come with strong information security?

    \item What is the price of \emph{not} protecting the information?
      \begin{itemize}
        \item Leaked company secrets.
        \item Inaccessible infrastructure.
      \end{itemize}

    \item Show incidents.

    \item Laws and other regulations?

\subsection{Project planning}

\begin{frame}{Project planning}
  \begin{itemize}
    \item MSB recommends that an \ac{isms} should start in project form and
      then go over to become a process.
  \end{itemize}
  \begin{figure}
    \includegraphics[height=0.5\textheight]{lis.png}
    \caption{Establish and implement an \ac{isms}.}
  \end{figure}
\end{frame}

\begin{frame}{Project plan}
  \begin{itemize}
    \item The foundation of the project.
    \item Defines the extent of the \ac{isms}\@.
    \item An agreement that is good for both the project leader and the
      management.
    \item Can contain:
      \begin{itemize}
        \item Background and need,
        \item purpose,
        \item goal,
        \item extent och demarcation,
        \item connections and contact surfaces,
        \item time plan, and
        \item budget.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Organising the project}
  \begin{itemize}
    \item Important with wide knowledge: to represent the entire organisation.
    \item Important with the correct qualifications: leading projects, and
      information security.
    \item Important with mandates!
    \item The organisation should be active and engaged, so that the competence
      is not gone once the project is over.
  \end{itemize}
\end{frame}

\begin{frame}{A short checklist}
  \begin{itemize}
    \item Have the management made a decision to implement \ac{isms}?
    \item Have the management appointed someone to coordinate the organisations
      work with information security?
    \item Have the management ensured that there is a strategy to communicate
      the work with \ac{isms} internally?
    \item Have the management made a decision regarding the budget and
      resources?
  \end{itemize}
\end{frame}

\section{Analyze}

\subsection{Organisational analysis}

\begin{frame}{What should be protected?}
  \begin{itemize}
    \item What information assets do we have, and are they worth protecting?
    \item What does it mean to protect them?
  \end{itemize}
\end{frame}

\begin{frame}{Organisational analysis}
  \begin{itemize}
    \item The purpose of the organisational analysis is to identify the
      informational assets available, and

    \item find how much they are worth protecting.

    \item Should lead to a structured list over
      \begin{itemize}
        \item what informational assets there are,
        \item what requirements and expectations they have, and 
        \item the worth of each asset.
      \end{itemize}

\begin{frame}{Example of informational assets}
  \begin{itemize}
    \item Employees: qualifications and experiences.
    \item Data: databases, agreement, documentation, samples, routines.
    \item Access to software: application software, system software, 
      development software.
    \item Services: data- and communication systems, supply systems.
    \item Immaterial: reputation, profile.
    \item Physical: computer equipment, movable media.
  \end{itemize}
\end{frame}

\begin{frame}{Dividing informational assets}
  \begin{description}
    \item[Primary] Refers to the main information, such as blue prints, logs or
      contracts.

    \item[Secondary] Refers to resources that are required to access the primary
      informational assets.

  \end{description}
  What will happen if you do not have your program that is required to read the
  closed proprietary format that the information is stored with?

\begin{frame}{Finding the informational assets}
  \begin{itemize}
    \item Previous process mappings?
    \item Department wise?
    \item IT-system?
    \item Project?
    \item Processes?
    \item By function?
  \end{itemize}
\end{frame}

\begin{frame}{Requirements on the informational assets}
  \begin{itemize}
    \item In order to be able to classify, the requirements needs to be known.

    \item Necessary to be able to objectively measure the need for protection.
      \begin{itemize}
        \item My information is the most important!
      \end{itemize}

    \item What information is necessary for the organisation?

\begin{frame}{Legal requirements}
  Agreement, laws and regulations.
  \begin{itemize}
    \item Personal Data Act,
    \item General Data Protection Regulation
    \item The Archives Act,
    \item Public Access to Information and Secrecy Act
    \item MSBFS 2016:1,
    \item Security Act.
  \end{itemize}
\end{frame}

\begin{frame}{Internal requirements}
  Requirements that the organisation have to reach its goals.
  For example:
  \begin{itemize}
    \item Vision,
    \item Business concept,
    \item policies,
    \item values.
  \end{itemize}
\end{frame}

@@

\begin{frame}{MSB:s suggestion on a classification model}
  \begin{figure}
    \includegraphics[height=0.7\textheight]{msb-klassificering.png}
    \caption{MSB:s suggestion on a classification model with levels: Severe, Considerable,
      Moderate, Negligible.}
  \end{figure}
\end{frame}

@@

\begin{frame}{University's adaptation of a classification model.}
  \begin{figure}
    \includegraphics[width=\textwidth]{miun-klassificering.png}
    \caption{University's adaptation of a classification model from the
      confidentiality perspective.}
  \end{figure}
\end{frame}

\begin{frame}{University's adaptation of a classification model}
  \begin{figure}
    \includegraphics[width=\textwidth]{miun-tillganglighet.png}
    \caption{University's adaptation of a classification model from the
      availability perspective.}
  \end{figure}
\end{frame}

\begin{frame}{University's adaptation of a classification model}
  \begin{figure}
    \includegraphics[width=\textwidth]{miun-riktighet.png}
    \caption{University's adaptation of a classification model from the
      integrity perspective.}
  \end{figure}
\end{frame}

\begin{frame}{Example of the result from the university}
  \begin{figure}
    \includegraphics[width=\textwidth]{miun-klassresultat.png}
    \caption{Part of the result from the university.}
  \end{figure}
\end{frame}

\subsection{Risk analysis}

\begin{frame}{Risk analysis}
  Necessary for demarcation:
  \begin{itemize}
    \item What informational assets should we do a risk analysis on?
    \item What informational assets are not critical enough for a risk analysis?
  \end{itemize}
\end{frame}

\begin{frame}{Risk analysis}
  \begin{itemize}
    \item Used to adapt the protection based on the assets of the organisation.
    \item Generate a list over
      \begin{itemize}
        \item existing threats,
        \item consequences of the threats, and
        \item suggestions for risk management.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Identify threats}
  \begin{itemize}
    \item Use brain storming to find potential threats.
    \item Include all the suggestions!
    \item Be specific: Intensional or unintentional information leakage.
    \item Example:
      \begin{itemize}
        \item An employee intentionally sabotages a system.
        \item An employee accidentally trips on a network cable.
        \item Software bug.
        \item Fire, flooding.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Risk matrix}
  \begin{figure}
    \includegraphics[height=0.7\textheight]{riskmatris.png}
    \caption{A risk matrix.}
  \end{figure}
\end{frame}

\begin{frame}{Risk matrix}
  \begin{itemize}
    \item Assess the probability and the consequence of each individual threat,
      and place them in the matrix.
    \item The focus on consequences should be consequences for the organisation.
    \item Gives a visual result, easily understandable and gives a good overview
      of how to prioritize security measures.
  \end{itemize}
\end{frame}

\begin{frame}{Risk matrix}{Consequences}
  \begin{itemize}
    \item How severe is the consequence for the organisation if the threat
      happens?
    \item Clarify who will be a affected
    \item Makes it easier to give examples for the different consequence levels.
      For example: loss of reputation, higher cost, \dots.
  \end{itemize}
  \begin{center}
    Severe -- Considerable -- Moderate -- Negligible
  \end{center}
\end{frame}

\begin{frame}{Risk matrix}{Probability}
  \begin{itemize}
    \item What is the probability that the threat occur?
    \item Makes it easier to give examples for the different levels: Years,
      weeks, days?
  \end{itemize}
  \begin{center}
    Very rarely -- Rare -- Regularly -- Often
  \end{center}
\end{frame}

\begin{frame}{Risk management}
  \begin{itemize}
    \item Decide whether the identified threats should be rectified or be
      accepted:
      \begin{itemize}
        \item Accept,
        \item Eliminate,
        \item Transfer,
        \item Rectify.
      \end{itemize}
    \item What measures should be taken?
  \end{itemize}
\end{frame}

>>>>

%\documentclass[handout]{beamer}
\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english,swedish]{babel}
\usepackage{booktabs}
\usepackage[iso,swedish]{isodate}
\usepackage{tabularx}
\usepackage{url}
\usepackage{varioref,prettyref}
\usepackage{graphicx}
\usepackage{color}
\usepackage{subfig}
\usepackage[style=alphabetic,maxbibnames=99,backend=bibtex]{biblatex}
\setbeamertemplate{bibliography item}[text]
\usepackage[varioref,prettyref]{miunmisc}

\addbibresource{literature.bib}

\mode<presentation>{
  \usetheme{Boadilla}
  \usecolortheme{}
  \usefonttheme{professionalfonts}
  \setbeamercovered{transparent}
}
\setbeamerfont{frametitle}{size=\footnotesize}
\setbeamerfont{framesubtitle}{size=\tiny}

\title[Intro infosäk]{%
  Introduction to\\
  Swedish Civil Contingencies Agency (MSB) methodological support for introducing \ac{isms}.
}
\author{Lennart Franked\footnote{Based on material written by
    Carina Bengtsson and Daniel Bosk}}
\institute[MIUN IST]{%
  Department of Informationsystem and Technologies (IST),\\
  Mid Sweden University, Sundsvall.

  %Avdelningen för informations- och kommunikationssytem (IKS),\\
  %Mittuniversitetet, Sundsvall.
}

\date{\today}

\pgfdeclareimage[height=0.65cm]{university-logo}{MU_logotyp_int_CMYK.pdf}
\logo{\pgfuseimage{university-logo}}

@@

\section[Metodstöd]{MSB:s metodstöd}

\begin{frame}{MSB}
  \begin{itemize}
    \item Swedish Civil Contingencies Agency (MSB).
    \item \enquote{MSB is responsible for issues concerning civil protection,
        public safety, emergency management and civil defence, as long as no
        other authority has responsibility\cite[About MSB]{msbse}\@.
        }
    \item This includes protection of our information systems.
  \end{itemize}
\end{frame}

\begin{frame}
  %Lista exempel, typ 1177(patientdata),
  %transportstyrelsen(registreringsinformation), Arbetsförmedlingen(felaktig
  %utbetalning).
\end{frame}

\begin{frame}{What is information security?}
  \begin{itemize}
    \item Occurs together with other processes and organisations.
    \item Information security in an organisation does not have a value by
      itself. It needs to be integrated into the organisation to be effective.
  \end{itemize}
\end{frame}

\begin{frame}{What is information security?}
  \begin{itemize}
    \item Ability to preserve the requirements and expectations that exist on
      information in an organisation.
    \item Amongst other to protect towards disruptions, such as the previously
      mentioned examples.
  \end{itemize}
\end{frame}

\begin{frame}{Demands and expectations on information}
  \begin{description}
    \item[Traceability] Who have taken part of, or changed the information?
    \item[Non Repudiation] It should not be possible to deny an act.
    \item[Authentication] Establish an entities identity.
    \item[Authorization] To give an authenticated entity certain permissions.
  \end{description}
\end{frame}

\begin{frame}{Why protect the information?}{Non-mandatory}
  \enquote{Good for business}
  \begin{description}
    \item[Reputation:] Who will let a company handle their information, if the
      company is known for treating their data carelessly.
    \item[Financial:] Strong reputation is better for the economy, and the cost
      of dealing with security incidents will be less.
    \item[Internal efficiency:] No loss of information or disruptions in the
      work.
    \item[Quality:] This will hopefully lead to a increase in work quality.
  \end{description}
\end{frame}

\begin{frame}{Governance documents}
  \begin{itemize}
    \item Plan of action, policies, guidelines, \dots
    \item Policy: What do we want to achieve?
      One extensive, or multiple smaller ones.
    \item The governance documents should be integrated into the organisations
      current structure.
  \end{itemize}
\end{frame}

\begin{frame}{Create commitment}
  \begin{itemize}
    \item The top management have the overall responsibility for the
      organisation -- This involves information security and security incidents.
    \item Important to make the management understand the importance of
      information security.
  \end{itemize}
\end{frame}

\begin{frame}{Motivation}
  \begin{itemize}
    \item What positive effects might come with strong information security?
    \item What is the price of \emph{not} protecting the information?
      \begin{itemize}
        \item Leaked company secrets.
        \item Inaccessible infrastructure.
      \end{itemize}
    \item Show incidents.
    \item Laws and other regulations?
  \end{itemize}
\end{frame}

\section{ISO 27000}
\begin{frame}{ISO 27001}
  \begin{itemize}
    \item ISO 27001 is a continuous process that strives towards constant
      improvements on how to work and use different security solutions in
      regards to information security.
    \item It is important to adapt this after the organisation, but this doesn't
      mean that parts can be skipped.
  \end{itemize}
\end{frame}

@@

@@

\begin{frame}{Methodological overview}
  \begin{figure}
    \includegraphics[width=6cm]{Figures/kugghjulet_lis.png}
    \caption{Overview over the methodological support~\cite{msb_metodstod}.}
  \end{figure}
\end{frame}

\begin{frame}{Different parts of the Methodological support}
  \begin{figure}
    \includegraphics[width=\textwidth]{Figures/oversiktsbild-over-metodstegens-olika-delar-och-hur-de-relaterar-till-varandra.png}
    \caption{Different parts of the Methodological support~\cite{msb_metodstod}}
  \end{figure}
\end{frame}

\begin{frame}{ISO 27000}
  \begin{itemize}
    \item ISO 27000 is an international standard covering working with
      information security.
    \item This is not just for governmental agencies.
    \item Any organisation can certify themselves for ISO 27000
    \item MSB:s methodological support is adapted to the international
      standards.
  \end{itemize}
\end{frame}

\section{Identify and Analyse}
\begin{frame}{Identify and Analyse --- Introduction}
  \begin{itemize}
    \item Analyse the external contexts (business intelligence)~\footnote{Omvärldsanalys} and the organisation.
    \item Requires knowledge about how the current situation looks
      like~\footnote{Nulägesanalys}.
      \begin{itemize}
        \item Look over existing documentation.
        \item Discuss within the organisation.
      \end{itemize}
    \item Mapping of the informational assets\footnote{informationskartläggning}.
    \item Analyse the risks connected to the informational assets.
    \item Find suitable security measurements.
    \item Existing security measurements are controlled using a GAP-analysis
  \end{itemize}
\end{frame}

\subsection{Business Intelligence (Omvärldsanalys)}
\begin{frame}{Identify and Analyse --- Business Intelligence}
  \begin{itemize}
    \item Mapping 
      \begin{itemize}
        \item External parties needs and expectations
          \begin{itemize}
            \item Customers
            \item Suppliers
            \item Citizens
            \item Reviewers
          \end{itemize}
        \item Prerequisites
          \begin{itemize}
            \item Technical
            \item Social
            \item Environmental
            \item Political
          \end{itemize}
      \end{itemize}
  \end{itemize}
\end{frame}
\begin{frame}{Example of external parties}
  \begin{table}
    \small
    \caption{Example of external parties~\cite{msb_metodstod}}
    \begin{tabularx}{\textwidth}{X X X}
      Category & Party & Requirement, Role or Impact\\
      \toprule
      Suppliers: &
      \begin{itemize}
        \item Suppliers
        \item Sub-suppliers
        \item Outsourcing partners
        \item In sourcing partners
        \item Cloud Service Providers
        \item Business Partners
        \item Security Providers
      \end{itemize} &
      An organisation is dependent on suppliers, can for example require \newline
      how the supplier is working with information security.\\
    \end{tabularx}
  \end{table}
\end{frame}

\begin{frame}{Legal Requirements --- Business Intelligence}
  \begin{description}
    \item[General Data Protection Regulation (GDPR)] Managing personal data.

    \item[Public Access to Information and Secrecy Act 2009:40] Says that
      certain information \emph{must} be available for the public, while other
      information should \emph{not} be.

    \item[The Archives Act 1990:782] Controls what documents that the
      authorities should preserve, that they should be ordered and looked after.
			It also controls how the documents can be disposed. Documents can not be
			disposed if that hinders the right to be able to provide access to official
			documents, research purposes or legal proceedings.
    \item[MSBFS 2016:1] Regulations regarding how governmental agencies should
      work with information security.
  \end{description}
  \begin{block}{Remember!}
    When going through the legal requirements, it is beneficial to get judicial
    expertise.
  \end{block}
\end{frame}
\begin{frame}{MSBFS 2016:1}
  \begin{itemize}
    \item Due to increased electronic information exchange in the society, there
      is now demands put on how governmental agency work with information
      security.
    \item The code of statutes came into effect \printdate{2010-02-01}.
  \end{itemize}
\end{frame}

\begin{frame}{GDPR}{General Data Protection Regulation}
  \begin{itemize}
    \item Replacing Personal Data Act 1998:204
    \item Regulates how personal data are allowed to be stored and used.
      \begin{itemize}
        \item Protects a persons right of privacy\cite[art. 5]{eu:gdpr}
        \item Personal data can only be stored for as long as there is a need
          for it\cite[art. 5]{eu:gdpr}.
        \item Each person must give explicit consent to a company/organisation
          to store, publish or share personal data about them\cite[art.
          7]{eu:gdpr}.  
        \item Ensure that a person have the right to get access
          to all personal
          data that has been collected about them\cite[art. 15]{eu:gdpr}.
        \item Persons right to erasure\cite[art. 17]{eu:gdpr}. 
      \end{itemize}
  \end{itemize}
\end{frame}
\begin{frame}{GDPR cont.}{General Data Protection Regulation}
    \begin{itemize}
      \item Possibility to transfer personal data (data
          portability)\cite[art. 20]{eu:gdpr}.  
      \item Data protection by design and default \cite[art. 25]{eu:gdpr}.
      \item Records regarding processing activities must be maintained \cite[art.
        30]{eu:gdpr}.
      \item Personal data breaches must be reported to the supervisory
        authority, without undue delay, not later than 72 hours\cite[art.
        33]{eu:gdpr}.
    \end{itemize}
\end{frame}
\begin{frame}{GDPR cont.}{General Data Protection Regulation}
\begin{itemize}
    \item Individuals needs to be informed about the personal data
        breach\cite[art. 34]{eu:gdpr}.
    \item An independent agency should ensure that these rules are followed
      \cite[art 51]{eu:gdpr}.
    \item If the organisation does not comply to these regulations, they can be
      fined up to 4\% worldwide turnover or €20 Million, whichever is greater
      \cite[art 83]{eu:gdpr}.
    \item and much more.
\end{itemize}
\end{frame}

\begin{frame}{MSBFS 2016:1}
  1 §  Denna författning innehåller föreskrifter som ansluter till
  bestämmelserna om statliga myndigheters informationssäkerhet i 19§
  förordningen (2015:1052) om krisberedskap och bevakningsansvariga
  myndigheters åtgärder vid höjd beredskap. 
\end{frame}

@@

  Detta innebär bland annat att en myndighet måste:
  \begin{enumerate}
    \item upprätta en informationssäkerhetspolicy och andra styrande dokument
      som behövs för myndighetens informationssäkerhet,
    \item utse en eller flera personer som leder och samordnar arbetet med
      informationssäkerhet,
    \item klassificera sin information med utgångspunkt i krav på
      konfidentialitet, riktighet och tillgänglighet,
    \item utifrån risk- och sårbarhetsanalyser och inträffade incidenter
      avgöra hur risker ska hanteras, samt besl uta om åtgärder för
      myndighetens informationssäkerhet,
    \item dokumentera granskningar och säkerhetsåtgärder av större betydelse
      som har vidtagits.
    \item
  \end{enumerate}
\end{frame}

@@

\subsection{Organisational Analysis\footnote{Verksamhetsanalys}}
\begin{frame}{Identify and Analyse --- Organisational Analysis}
  \begin{itemize}
    \item Mapping
    \begin{itemize}
      \item Internal parties needs, requirements and expectations.
      \begin{itemize}
        \item Decision makers
        \item Object owners
        \item Co-workers
        \item Support units
      \end{itemize}
      \item Prerequisites.
      \begin{itemize}
        \item Goals
        \item Strategies
        \item Organisational Culture
        \item Infrastructure
      \end{itemize}
      \item Identifies the organisations essential informational assets.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Analyse Informational Assets}
  \begin{block}
    Informational assets are the organisations information and its resources
    that deals with information. For example to receive, store, process, show or
    communicate it~\cite{msb_metodstod}.
  \end{block}
\end{frame}

\begin{frame}{Dividing informational assets}
  \begin{description}
    \item[Primary] Refers to the main information, such as blue prints, logs or
      contracts.
    \item[Secondary] Refers to resources that are required to access the primary
      informational assets.
  \end{description}
  What will happen if you do not have your program that is required to read the
  closed proprietary format that the information is stored with?
\end{frame}

\begin{frame}{Organisational Analysis}
  \begin{block}{General analysis}
    Important to only perform a general analysis of the organisation.
  \end{block}
  \begin{block}{Previous Work?}
    Are there any previous analysis available?
  \end{block}
\end{frame}

@@

\begin{frame}{MSB:s suggestion on a classification model}
  \begin{figure}
    \includegraphics[height=0.7\textheight]{Figures/msb-klassificering.png}
    \caption{MSB:s suggestion on a classification model with levels: Severe, Considerable,
      Moderate, Negligible.}
  \end{figure}
\end{frame}

@@

\begin{frame}{University's adaptation of a classification model.}
  \begin{figure}
    \includegraphics[width=\textwidth]{Figures/miun-klassificering.png}
    \caption{University's adaptation of a classification model from the
      confidentiality perspective.}
  \end{figure}
\end{frame}

\begin{frame}{University's adaptation of a classification model}
  \begin{figure}
    \includegraphics[width=\textwidth]{Figures/miun-tillganglighet.png}
    \caption{University's adaptation of a classification model from the
      availability perspective.}
  \end{figure}
\end{frame}

\begin{frame}{University's adaptation of a classification model}
  \begin{figure}
    \includegraphics[width=\textwidth]{Figures/miun-riktighet.png}
    \caption{University's adaptation of a classification model from the
      integrity perspective.}
  \end{figure}
\end{frame}

\begin{frame}{Example of the result from the university}
  \begin{figure}
    \includegraphics[width=\textwidth]{Figures/miun-klassresultat.png}
    \caption{Part of the result from the university.}
  \end{figure}
\end{frame}

\subsection{Risk analysis}
\begin{frame}{Risk analysis}
  \begin{itemize}
    \item Identifies the information security risks.
    \item Can be performed over generally over the organisation, or on a
      specific object.
    \item The identified risks are then assessed based on probability and
      consequence.
  \end{itemize}
\end{frame}

\begin{frame}{When should the Risk Analysis be performed?}
  \begin{itemize}
    \item Not necessary to perform the risk analysis after the business
      intelligence and organisational analysis.
    \item If it is performed after, then the results from these can be used in
      the risk analysis.
    \item Otherwise, it is possible to rely on the existing knowledge of the
      organisation.
  \end{itemize}
\end{frame}

\begin{frame}{How should the process for Risk Analysis look like?}
  \begin{block}{Already have a process?}
    If the organisation already have a process in place, then this can be used.
    Since it's already well established in the organisation.
  \end{block} 
  \begin{block}{If there isn't a process in place?}
    Some key points to think about when establishing a process for Risk
    Analysis.
    \begin{itemize}
      \item How should the process look like?
      \item What criteria for consequence and probability estimation should be
        used?
      \item What criteria for risk acceptance should be used?
      \end{itemize}
      In IS0 27001:2018 the management are responsible for making a decision
      about the process, criteria for consequence and probability levels, and
      what risk levels should be considered acceptable. 
  \end{block}
\end{frame}

\subsection{GAP Analysis}
\begin{frame}{GAP Analysis}
  \begin{itemize}
    \item Identifies the gap between the current state of the organisation and
      what level of security the organisation wants to achieve.
    \item For example, the current state versus the actions mentioned in ISO
      27001~\cite[Bilaga A]{iso27001}
    \item Develop an action plan on how to fix the objects that aren't 
      sufficient.
  \end{itemize}
\end{frame}

\subsection{The Results of the Analysis}
\begin{frame}{The result of the analysis}
  \begin{itemize}
    \item The result will act as a base when modelling the information security.
    \item Helps to assess how the organisation handles the information security.
    \item Identifying security measures.
  \end{itemize}
